---
title: "Lab 1"
author: "Olivia Roberts"
format: html
embed-resources: true
fig-width: 5
fig-height: 4
theme: "cyborg"
---

```{r}

library(datasauRus)
library(ggplot2)

```

**Question 1**

```{r}

?datasaurus_dozen

table(datasaurus_dozen$dataset)

```

Answer: The datasaurus_dozen file contains 1846 rows of 3 variables. The 3 variables are a variable called "dataset" that indicates the dataset the data is from, and x-values variable, and a y-values variable. By accessing the dataset column, we can see a list of the 13 datasets inclued in datasaurus_dozen and that they have 142 rows each.

**Question 2**

```{r}

dino_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'dino',]

ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()

cor(dino_data$x, dino_data$y)

```

**Question 3**

```{r}

star_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'star',]

ggplot(data = star_data, mapping = aes(x = x, y = y)) +
  geom_point()

cor(star_data$x, star_data$y)

```

Answer: The correlation coefficient for the dino and star datasets are very similar; they are -0.064 and -0.063 respectively. They are both negative, so in both datasets, as x increases, y decreases. The dino dataset has a slightly stronger negative correlation, but neither of the datasets have a very strong negative correlation since the points are very scattered and the magnitude of the correlation is not very large. Despite having similar correlations, the two datasets look very different.

**Question 4**

```{r}

circle_data <- datasaurus_dozen[datasaurus_dozen$dataset == 'circle',]

ggplot(data = circle_data, mapping = aes(x = x, y = y)) +
  geom_point()

cor(circle_data$x, circle_data$y)

```

Answer: The correlation coefficient of the dino dataset is -0.064, while the correlation coefficient for the circle dataset is -0.068. Again, both datasets have a negative correlation (as x increases, y decreases overall), but the circle dataset now has a slightly stronger magnitude of correlation. However, both datasets still have a very similar magnitude of correlation despite looking very different.

**Question 5**

```{r}
#| fig-height: 12
#| fig-width: 8

ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")

```

**Question 6**

```{r}

sapply(unique(datasaurus_dozen$dataset), function(name){
    subset <- datasaurus_dozen[datasaurus_dozen$dataset == name, ]
    return(cor(subset$x, subset$y))
})

```
